% Init .tex document for a collections of questions of the VCS exam

\documentclass[12pt]{article}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsfonts}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{listings}
\usepackage{color}
\usepackage{float}
\usepackage{subcaption}
\usepackage{enumitem}


\title{VCS Exam Questions}
\author{Marco Bernardi}
\date{\today}

\begin{document}
\maketitle

\section{Introduction}
This document contains a collection of questions that can be used to prepare for the VCS exam.
\section{Machine Learning}
\begin {enumerate}
\item \textbf{Describe the main components of a supervised machine learning model (e.g. linear regression),
    i.e. the hypothesis function, the cost function and how the learning works.}


\end{enumerate}

\section{Filters}
\begin{enumerate}
    % Write in bold the question 
    \item \textbf{Please provide a clear definition for a Gaussian filter (i.e. the kernel mask) and introduce another filter that can give you a similar output.
              You should also describe the main differences between the two.}

          Gaussian filter is a smoothing filter that is used to reduce noise in an image by removing high-frequency components.
          The kernel mask of a it is an approximation of a 2D Gaussian function in which pixels closer to the center have higher weights than pixels further away.

          The filter has two main parameters:
          \begin{itemize}
              \item $\sigma$: the standard deviation of the Gaussian distribution, which controls the amount of smoothing applied to the image.
                    Increasing $\sigma$ means that more pixels are included in the smoothing operation, resulting in a more blurred image while decreasing $\sigma$ results in less smoothing and more detail in the image.
              \item $k$: the size of the kernel mask, which determines the extent of the smoothing operation.
                    A larger kernel approximates the Gaussian function more closely but requires more computation.
          \end{itemize}

          The sum of the elements in window is equal to 1, because constant values in the image are preserved.

          \[\frac{1}{16}
              \begin{bmatrix}
                  1 & 2 & 1 \\
                  2 & 4 & 2 \\
                  1 & 2 & 1
              \end{bmatrix}
              = h(u, v) = \frac{1}{2\pi\sigma^2}e^{-\frac{u^2 + v^2}{2\sigma^2}}
          \]

          Another filter that can give a similar output is the box filter, which is a simple averaging filter that replaces each pixel in the image with the average of its neighbors.
          The box filter has a constant kernel mask in which all elements are equal, and the sum of the elements in the kernel is equal to 1.

          \[\frac{1}{9}
              \begin{bmatrix}
                  1 & 1 & 1 \\
                  1 & 1 & 1 \\
                  1 & 1 & 1
              \end{bmatrix}
          \]

          The main difference between the Gaussian filter and the box filter is that the Gaussian filter gives more weight to pixels closer to the center of the kernel, while the box filter gives equal weight to all pixels.

    \item \textbf{You have this kernel:     16 * [1, 2, 1; 2, 4, 2; 1, 2, 1]
              Define what is it and another kernel with which you can achieve a similar result.}

          The kernel is a Gaussian filter with $\sigma = 1$ and $k = 3$.
          The kernel approximates a 2D Gaussian function in which pixels closer to the center have higher weights than pixels further away.
          The sum of the elements in the kernel is equal to 16, because the kernel is normalized to preserve the constant values in the image.

          Another kernel that can achieve a similar result is the box filter, which is a simple averaging filter that replaces each pixel in the image with the average of its neighbors.
          The box filter has a constant kernel mask in which all elements are equal, and the sum of the elements in the kernel is equal to 1.


    \item \textbf{What’s the purpose of image filtering?  You should give a clear definition of filtering and also define the mathematical tools used in order to define any filter.
              Finally, give an example of a smoothing filter (and provide also its definition).}


          Image filtering is a process used in image processing to enhance, modify, or extract features from an image by manipulating pixel values. The purpose of filtering is to remove noise, detect edges, smoothen, sharpen, or emphasize certain features in an image.

          Filters are defined using mathematical operations applied to pixel neighborhoods (small groups of adjacent pixels) in an image. Common mathematical tools include:

          \begin{itemize}
              \item \textbf{Convolution:} A filter (or kernel) is slid over the image, and pixel values are modified based on a weighted sum of their surrounding pixels. Convolution is mathematically expressed as:
                    \[
                        I'(x, y) = \sum_{i=-k}^{k} \sum_{j=-k}^{k} K(i, j) \cdot I(x-i, y-j)
                    \]
                    where:
                    \begin{itemize}
                        \item \(I(x, y)\) is the original image,
                        \item \(I'(x, y)\) is the filtered image,
                        \item \(K(i, j)\) is the filter kernel (a matrix of weights),
                        \item \(k\) defines the kernel size.
                    \end{itemize}

              \item \textbf{Fourier Transform:} Filters can also be applied in the frequency domain by transforming the image using a Fourier Transform, modifying its frequency components, and then transforming it back to the spatial domain.

              \item \textbf{Non-linear Operations:} Some filters, such as median or bilateral filters, involve non-linear operations where pixel values are replaced based on specific rules instead of simple arithmetic.
          \end{itemize}

          A smoothing filter reduces noise and smoothens the image by averaging the pixel values in a neighborhood.

          A common example of a smoothing filter is the \textbf{box filter (uniform average filter)}, which averages the values in a fixed-size neighborhood equally. Its kernel is:
          \[
              K = \frac{1}{n^2}
              \begin{bmatrix}
                  1 & 1 & 1 \\
                  1 & 1 & 1 \\
                  1 & 1 & 1
              \end{bmatrix}
          \]
          For a \(3 \times 3\) kernel (\(n=3\)), the output at a pixel is:
          \[
              I'(x, y) = \frac{1}{9} \sum_{i=-1}^{1} \sum_{j=-1}^{1} I(x+i, y+j)
          \]
          This filter smoothens the image by averaging pixel values over a \(3 \times 3\) grid, reducing abrupt changes.

    \item \textbf{Please provide i) a clear definition of the convolutional operator, ii) its main propertiesand iii)
              report the kernel matrix of a filter that will produce as an output an image that is shifted by 2 pixels in the left direction.}

          Convolutional operator is a mathematical operation that takes two functions f and g and produces a third function that represents how the shape of one is modified by the other.
          The convolution of f and g is written f * g. The convolution operator is often seen in signal processing, where it models the effect of a linear time-invariant system on a signal.
          Convolutional operator follows the classical mathematical properties of associativity, commutativity, and distributivity, homogeneity, shift invariance, and separability.
          In image processing, the convolution operator is used to apply filters to images.
          The kernel matrix of a filter that will produce as an output an image that is shifted by 2 pixels in the left direction is:
          \[
              \begin{bmatrix}
                  0 & 0 & 0 \\
                  0 & 0 & 2 \\
                  0 & 0 & 0
              \end{bmatrix}
          \]

    \item \textbf{What is this filter [1 0 -1; 2 0 -2; 1 0 -1]?  Please “decompose” it in its two 1D components and highlight their respective roles.}

          The following filter is a Sobel filter applied to vertical edges. The filter can be decomposed into two 1D components:
          \[
              \begin{bmatrix}
                  1 \\
                  2 \\
                  1
              \end{bmatrix}
              \begin{bmatrix}
                  1 & 0 & -1
              \end{bmatrix}
          \]
          The first component is a weighted average and scaling filter that emphasizes the vertical edges in the image.
          It's important to use a smoothing filter to reduce the noise in the image before applying the Sobel filter, because derivatives are sensitive to noise.
          The second component is a derivative filter that computes the gradient of the image in the vertical direction.
          The combination of the two filters highlights the vertical edges in the image.

          Sobel filter can be used also for horizontal edges by swapping the two components of the filter.
\end{enumerate}

\section{Edge}

\begin{enumerate}
    % Write in bold the question 
    \item \textbf{Describe an algorithm for edge detection (and define the kernel used). Report eventual similarities between this procedure and human vision.}

    \item \textbf{Please provide an example of edge detector (i.e. the kernel matrix) and illustrate whatare the main analogies with biological vision (if any),
              and the applications in which edgedetection could be a fundamental component.}

    \item \textbf{Describe in detail the SIFT algorithm; you should present both detection and description stages,
              and illustrate the classical SIFT matching approach (aka Lowe’s 2NN ratio test).}

    \item \textbf{Describe all the major components of the SIFT algorithm and provide some examples ofapplications in which SIFT features are the key ingredient.}

          Ciao

\end{enumerate}

\section{BOVW}
\begin{enumerate}
    % Write in bold the question 
    \item \textbf{Describe in detail a bag-of-visual-words model by highlighting all the main steps of computation and the major “design choices” /parameters.
              Then you should briefly introducespatial pyramids and motivate their advantages with respect to a standard pipeline.}

    \item \textbf{Describe in details a bag-of-visual-words pipeline. What are its main stages and what are their roles?}

          Ciao
    \item \textbf{Describe the Bag of Word pipeline. Propose a technique to address feature locality.}

    \item \textbf{Describe in details a bag-of-visual-words pipeline highlighting all the main steps of com-putation.
              Finally, you should briefly introduce how it can be extended in order to betterrepresent the spatial elements depicted in a given image.}

\end{enumerate}

\section{CNN}
\begin{enumerate}
    % Write in bold the question 
    \item \textbf{Describe in detail a CNN architecture, introducing the most important components (layers, activation functions, etc.)
              and pros/cons with respect to a traditional neural network.}

    \item \textbf{Describe the main components of a CNN (conv, pooling and FC layers). Report a typical architecture (take example from AlexNet).}

    \item \textbf{What’s a perceptron? Describe in details its main components and how a multi-layerperceptron can add more representation power w.r.t. a simple perceptron.}

\end{enumerate}
\end{document}